---
title: "Merging Spatial Data"
author: "Nick Eubank, building off excellent tutorials by Claudia Engel"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: spacelab
    mathjax: default
    fig_width: 6
    fig_height: 6
---


```{r knitr_init, echo=FALSE, cache=FALSE, message=FALSE,results="hide"}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)

library(rgdal)
library(sp)
library(rgeos)
library(plyr)
library(raster)

pk.dist <- readOGR(dsn = "RGIS2_data/shapefiles", layer="pk_districts")
strikes <- readOGR(dsn = "RGIS2_data/shapefiles", layer="pk_drone_strikes")
dist.crs <-CRS(proj4string(pk.dist))
strikes.projected <- spTransform(strikes, dist.crs)
pk.pop <- raster("RGIS2_Data/pakp00g")
new.crs <- CRS(projection(pk.pop))
pk.dist.rasterproj <- spTransform(pk.dist, new.crs)


setwd("~/documents/polgeo/gis_in_r")
districts<-readOGR("RGIS2_Data/shapefiles",'congressional_districts')
grants<-readOGR("RGIS2_Data/shapefiles",'federal_grants')
newcrs<-CRS(proj4string(districts))
grants.newproj <- spTransform(grants, newcrs)


```

***

When working with spatial data, one is rarely interested in working with only one source of data. This tutorial will introduce a set of tools for linking vector data with other data sources. It begins by introducing how to link spatial vector data with non-spatial data from in table format, then turns to the problem of linking multiple sources of spatial data through spatial joins and intersects. 

This tutorial uses the `sp`, `rgdal`, and `raster` libraries from the `RGIS1` tutorial. If you have not yet installed those, please revisit that tutorial for directions. In addition, this tutorial will also make use of the `rgeos` library, installation of which is discussed in `part0_setup`. 

***

# 1. RGEOS Library

The `over` command will cover a remarkable number of situations, but it is really only interesting if two shapes intersect exactly. In practice, however, we are often interested in sligthly more flexible questions: how many cities are *within 10km* of a drone strike? Or how many people live close to a government project? 

When we want to do fancer geometric operations, we use the `rgeos` library. `rgeos` (which stands for "R interface to the *Geometry Engine - Open Source* library") is a set of tools for geometric operations. `GEOS` a huge library, and one that underlies many spatial tools (not just in R). Whenever you're thinking about a geometric operation, `rgeos` is the first place to look.


`rgeos` tools can be broadly divided into three camps. Here some of the most commonly used ones. 

### Calculating Properties

* **gArea**: calculate area of a shape
* **gLength**: calculates length of line / circumference of polygons
* **gDistance**: distance between items

### Making New Shapes

* **gBuffer**: Expand points into circles of given radius
* **gCentroid**: Collapse polygons to their centroids
* **gDifference**, **gUnion**, and **gIntersection**: execute set operations on polygons
* **gUnionCascaded**: dissolves a collection of shapes into a single shape. 
* **gSimplify**: If your polyon is too high a resolution (higher than needed for analysis, and too high for fast processing), reduces the number of vertices while maintaining shape to the best of it's ability. 


### Testing Geometric Relationships

* **gIntersects**: test if shapes intersect. Primarily useful for testing whether two polygons intersect, since this is not something`over` can do. 
* **gContains**: Is one spatial object entirely within another?
* **gIsValid**: Very useful -- make sure your geometries aren't corrupt!


## Example 1: Calculating Area

The syntax for these tools varies both across tools and depending on the input types, but here are some basic examples with the congressional district data. 

Area of districts:
(this is actually already in the districts data, but let's recompute anyway!)
```{r}
areas <- gArea(districts, byid=TRUE)
areas

districts@data <- cbind(districts@data, areas)
```

This simple example demonstrates two important things about `rgeos`: 

* **Units**: `rgeos` isn't a spatial library -- it just does geometry. It takes x-y coordinates and applies geometric formula. Thus the units will always be the units of the x-y coordinates, which come from your projection.
* **`byid`**: `byid` is an option on most `rgeos` commands. If `byid=TRUE`, each observation in a Spatial* object is handled separately; if `byid=FALSE`, a Spatial* object is treated as one big geometry. So if one were working with a `SpatialPolygons` object of US states, when `byid=TRUE`, the analysis would be conducted for each state; if `byid=FALSE`, it would essentially run the analysis against the United States as a whole. 

For example, if we didn't change `byid` to `TRUE`, we'd get a single measure of area for all the congressional districts together:

```{r}
gArea(districts, byid=TRUE)
```


## Example 2: Buffer

A common operation is to create buffers (polygons of fixed radius) around points. Let's make 7km buffers around government grants.

```{r}
summary(grants.newproj) # Check projection to make sure we know units!
buffered.grants <- gBuffer(grants.newproj, width=7000) # 5km = 7,000 meters. 

plot(districts)
par(new=T)
plot(buffered.grants, col='blue', add=T)
```

Note that because we were interested in areas within 5km of *any* government grant, we left `byid=FALSE` (the default), so `buffered.grants` has only 1 feature. If we had passed `byid=TRUE` we could have gotten different polygons (many overlapping) for each grant. 

## Example 3: Set operations

The figure above looks a little odd because the buffered polygons spread into the ocean. We can fix this by "clipping" the buffered polygons using `gIntersection`. 

```{r}
intersection <- gIntersection(buffered.grants, districts, byid=TRUE)
plot(districts)
par(new=T)
plot(intersection, col='blue', add=T)
par(new=F)
```

Interestingly, note that while `buffered.grants` was previously one feature, because we passed `byid=TRUE`, the intersection with the districts executed the union with each congressional district, creating three distinct polygons:

```{r}
intersection
```

However, note that `rgeos` struggles if we just change `byid` to `FALSE`. The following code:
```{r, eval=FALSE}
gIntersection(buffered.grants, districts, byid=FALSE)
```
Generates the error:
`Error in RGEOSBinTopoFunc(spgeom1, spgeom2, byid, id, drop_lower_td, "rgeos_intersection") :` 
`  TopologyException: no outgoing dirEdge found at 570113.57942913717 4148852.3204373871`

This kind of problem will come up from time to time. Remember that `rgeos` is working with objects whose vertices are defined by numbers of limited precision -- sometimes it'll have weird problems, and you have to experiment. In this case, I *think* this is because the edges of the congressional districts don't quite coincide due to problems of machine precision. To fix this, we can dissolve the districts first using `gUnionCascaded`:

```{r}
merged.districts <- gUnionCascaded(districts)
plot(merged.districts)

intersection2 <- gIntersection(buffered.grants, merged.districts, byid=FALSE)
intersection2
plot(districts)
par(new=T)
plot(intersection2, col='blue', add=T)
par(new=F)
```
Now the intersection is only one part!



### Exercise 1

Let's try and figure out what percentage of residents of Pakistans Federally Administered Tribal Areas (FATA, the regions with the most drone strikes) live within 1km of a drone strike. 

#. Subset the pk.dist data for the Province of Fata.
#. Create a buffer of 1km around each drone strike with the `gBuffer` command. 
#. Create a buffer around the center
#. Select all polys that fall in the buffer

Let's do it! 



***

**Answers to Exercise 1**


